Well-quasi-orderings on word languages

### 2025-10-15 src/amalgam.tex
fix: making knowledge work

\AP Our second focus is of practical nature: we want to give a decision
procedure for being \kl{well-quasi-ordered}. This will require us to introduce
\emph{effectiveness assumptions} on the \kl{amalgamation systems}. While most
of them will be innocuous, an important consequence is that we have to consider
\emph{classes of languages} rather than individual ones, for instance: the
class of all regular language, or the class of all context-free languages. Such
classes will be called \kl[effectiveness assumptions@amalgamation]{effective amalgamative classes} (\kcref{effectiveness assumptions@amalgamation})\todo{can't get the knowledge to work}. In the following theorem, we prove that under such
assumptions, testing \kl{well-quasi-ordering} is inter-reducible to testing
whether a language of the class is empty, which is usually the simplest
problem for a computational model.

@@

%For this paper to be self-contained, we
%will also recall how runs of a finite state automaton can be understood as an
%\kl{amalgamation system}.
%
%\begin{example}[{\cite[Section 3.2]{ASZZ24}}]
%    Let $A = (Q, \delta, q_0, F)$ be a finite state automaton over a finite
%    alphabet $\Sigma$. Let $\Delta$ be the set of transitions $(q_1, a, q_2)
%    \in Q \times \Sigma \times Q$,
%    and $R \subseteq \Delta^*$ be the set of 
%    words over transitions that start with the initial state $q_0$,
%    end in a final state $q_f \in F$, and such that the end state of a
%    letter is the start state of the following one.
%    The canonical decomposition $\canrun$
%    is defined as a morphism from $\Delta^*$ to $\Sigma^*$
%    that maps $(q,a,p)$ to $a$. 
%    Because of the one-to-one correspondence of steps of a run $\rho$ and letters in its \kl{canonical decomposition}, 
%    we may treat the two interchangeably.
%    Finally, given two runs $\rho$ and $\sigma$ of the automaton,
%    we say that an embedding $f \in \HigEmb(\canrun(\rho), \canrun(\sigma))$
%    belongs to $E(\rho,\sigma)$ when
%    $f$ is also defining an embedding from $\rho$ to $\sigma$ as words in $\Delta^*$.
%
%    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
%    whose language is precisely the language of words recognized
%    by the automaton $A$.
%\end{example}
%\begin{proof}
%    By definition, the embeddings inside $E(\rho,\sigma)$ are in
%    of $\HigEmb(\canrun(\rho), \canrun(\sigma))$, and they compose properly.
%    Because $\Delta = Q \times \Sigma \times Q$ is finite, it is 
%    a \kl{well-quasi-ordering} when equipped with the equality relation, and 
%    we conclude that $\Delta^*$ with $\higleq$ is a \kl{well-quasi-order}
%    according to Higman’s Lemma \cite{HIG52}.
%    
%    Let us now move to proving that the system satisfies the amalgamation
%    property. Given three runs $\rho,\sigma,\tau \in R$, and two embeddings $f \in E(\rho,\sigma)$
%    and $g \in E(\rho,\tau)$, we want to construct an amalgamated run $\sigma \vee \tau$.
%    Because letters in the run $\rho$ respect the transitions of the automaton
%    (i.e., if the letter $i$ ends in state $q$, then the letter $i+1$ starts in
%    state $q$), then the \kl{gap word} at position $i$ starts in state $q$ and
%    ends in state $q$ too. This means that for both embeddings
%    $f$ and $g$, the \kl{gap words} are read by the automaton by looping
%    on a state. In particular, these loops can be taken in any order
%    and continue to represent a valid run. That is, we can even select
%    the order of concatenation in the amalgamation for \emph{all} 
%    $0 \leq i \leq \card{\canrun(\rho)}$ and not just for one separately.
%
%    We conclude by remarking that 
%    the language of this amalgamation system is
%    the set of $\yieldrun(R)$, 
%    because $R$ is the set of valid runs of the automaton,
%    and $\yieldrun(\rho)$ is the word read along a run $\rho$.
%\end{proof}

@@

\AP To construct a decision procedure for well-quasi-orderedness under $\infleq$, we 
our \kl{amalgamation systems} to satisfy certain \intro{effectiveness assumptions}. We require that for an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$, $R$ is
recursively enumerable, the function $\canrun(\cdot)$ is computable, and for
any two runs $\rho, \sigma \in R$, the set $E(\rho,\sigma)$ is computable. Additionally, we require the class to be effectively closed under rational transductions \cite[Chapter 5, page 64]{BERST79}.

>>>>

\AP Our second focus is of practical nature: we want to give a decision
procedure for being \kl{well-quasi-ordered}. This will require us to introduce
\emph{effectiveness assumptions} on the \kl{amalgamation systems}. While most
of them will be innocuous, an important consequence is that we have to consider
\emph{classes of languages} rather than individual ones, for instance: the
class of all regular language, or the class of all context-free languages. Such
classes will be called \kl(amalgamation)[effectiveness assumptions]{effective
amalgamative classes} (\kcref{effective amalgamation system}). In the
following theorem, we prove that under such assumptions, testing
\kl{well-quasi-ordering} is inter-reducible to testing whether a language of
the class is empty, which is usually the simplest problem for a computational
model.

@@

@@

\AP To construct a decision procedure for well-quasi-orderedness under
$\infleq$, we our \kl{amalgamation systems} to satisfy certain
\intro(amalgamation){effectiveness assumptions}. We require that for an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$, $R$ is recursively
enumerable, the function $\canrun(\cdot)$ is computable, and for any two runs
$\rho, \sigma \in R$, the set $E(\rho,\sigma)$ is computable. Additionally, we
require the class to be effectively closed under rational transductions
\cite[Chapter 5, page 64]{BERST79}.
